
\chapter{From discrete to continuous quantum systems}\label{cap:results}




Quantum mechanics reveals nature in two seemingly different yet deeply connected forms: systems with a finite set of distinct states, and systems described by a continuous range of possibilities. In the simplest cases—such as spin systems or qubits—the structure is discrete, with states that can be counted and represented within finite-dimensional spaces. Yet many of the most fundamental physical systems, including particles moving in space or modes of electromagnetic fields, require a continuous description, where variables such as position and momentum take values across an unbroken spectrum.

Historically, the development of quantum theory began with the striking discovery that certain physical quantities, such as the energy levels of atoms, are quantized. These early successes naturally led to models built on discrete states. However, as the theory matured, it became clear that a complete description of quantum phenomena must also incorporate continuous degrees of freedom, captured by wavefunctions and operators acting on infinite-dimensional spaces. This shift did not merely extend the formalism—it reshaped the conceptual and mathematical framework of the theory.

The contrast between these two regimes can be illustrated by comparing a two-level system to a quantum particle in one dimension. While the former can be fully described by a small set of amplitudes, the latter requires a function defined over all space, encoding a superposition of infinitely many possible outcomes. Moving between these perspectives involves more than increasing complexity; it demands new tools, replacing finite vectors with functions, matrices with operators, and sums with integrals.

Understanding this transition is essential for engaging with a wide range of physical systems, from quantum optics to field theory. At the same time, it introduces subtle challenges, including the proper handling of normalization, the interpretation of measurements in continuous spectra, and the rigorous treatment of infinite-dimensional spaces.

This chapter explores how the principles established for discrete quantum systems extend to the continuous domain. How does one generalize familiar concepts such as states, observables, and measurements when the set of possibilities is no longer countable? And what new insights emerge when we move from isolated points to continuous structures? By addressing these questions, we build a bridge between two foundational perspectives of quantum theory.


This chapter discuss the thermal relaxation asymmetry for discrete and continuous-variable quantum systems models.

Therefore, studying both discrete and continuous systems enables the exploration of different physical and mathematical resources, as well as the construction of bridges between models that expand both the theoretical understanding and experimental possibilities of quantum theory.

%%%%%%

\section{Two Interacting Qubits}\label{sec:qubits}



\begin{equation}
    H_S = -\frac{\omega_0}{2} \sigma_{1}^{z} - \frac{\omega_0}{2} \sigma_{2}^{z} + g (\sigma_{1}^{+} \sigma_{2}^{-} + \sigma_{1}^{-} \sigma_{2}^{+}),
    \label{eq:HS-2qubits}
\end{equation}


\subsection{Thermal Relaxation of Two Interacting Qubits}\label{sec:result-qubits}


%%%%%%

\section{Gaussian States}\label{sec:gaussian}

\subsection{Thermal Relaxation of Gaussian States}\label{sec:result-gaussian}

\subsubsection{Asymmetry for Thermal States}\label{sec:result-thermal}

\subsubsection{Asymmetry for Displaced States}\label{sec:result-displaced}

\subsubsection{Asymmetry for Squeezed States}\label{sec:result-squeezed}