

%% USPSC-Cap3-Conclusao.tex
% ---
% Conclusão
% ---
\chapter{Conclusion}\label{cap:conclusion}

The main question investigated in this work was how quantum properties affect the heating-cooling asymmetry in two-interacting qubits and Gaussian systems. The relaxation dynamics was described by the Lindblad master equations and we used the thermal kinematics approach to identify the heating-cooling asymmetry. In this respect, we introduced an adaptation of thermal kinematics to Gaussian systems. In addition to thermal kinematics, we analysed how entropy production can be used to explain which thermal relaxation process is faster.

For the two-interacting-qubits model, we investigated the thermal relaxation of systems in the presence of global coherence and entanglement. By applying Davies maps to this model, the global state always thermalizes to the bath's temperature, but the interaction between the qubits can generate correlations between them, and, as a result, we cannot write the global steady state as the tensor product of qubit's thermal states. However, we can associate an effective temperature to the qubits with the cost of this temperature at equilibrium being lower than the bath's temperature. Another consequence of the qubits interaction generating correlations is that the local entropy production rate becomes negative at some periods, which indicates the a local non-Markovian dynamics. We noticed that the thermal relaxation case with non-Markovianity was faster than the no-interaction and full Markovian case. However, we observed that strong interaction causes entanglement, which reduces the asymmetry and make the relaxation slower than the weaker interaction without entanglement generation. Additionally, we saw that initial global coherence can apparently increase the asymmetry, changing the intermediary dynamics, but it do not affect the total equilibration time of the qubits.

For the continuous-variable case, we explored the heating-cooling asymmetry in single-mode Gaussian states, where the investigated resource was ergotropy that separate how much of the nonequilibrium energy is thermodynamically useful. Because the passive state of Gaussian systems is always a thermal state, any quantumness of the system will be related to ergotropic quantities. An interesting consequence of using the ergotropic approach is the possibility of separating the Wigner-Fisher information and the entropy production rate into passive and ergotropic contributions. So, we can separate the asymmetry origin from passive contributions and from ergotropic contribution.

The first Gaussian system we analysed was thermal states, in which we analytically showed that heating is always faster than cooling and where entropy production successfully explained its origin. Second, we compared the relaxation of different thermal states that were equally displaced and showed that displacement does not affect the asymmetry, being the passive contribution its generator. Again, the total entropy production explained heating being faster than cooling and separating the contributions showed that the ergotropy was the same for both heating and cooling, which is a consequence of being independent of the system's initial temperature. Then, we study the asymmetry in squeezed thermal states where both ergotropic and passive contributions are affected by the Gaussian operation. In this case, using bigger squeezing parameters resulted in a reduction in the asymmetry of the total state while the passive state had its asymmetry increased. To explain the reduction of the total state asymmetry, we observed that increasing the squeezing parameter increases of the ergotropic contribution but it is not enough to compensate the decrease of the passive contribution, resulting in a total entropy production decrease. Also, increasing the squeezing parameter also increased the difference between the heating and cooling passive state's initial entropy production, resulting in the asymmetry increase. A last result from the asymmetry with squeezed states it that we identified a non-Markovian dynamics for the heating passive state of the big squeezing state and we also observed an acceleration of the dynamics. As a result, once again we have an indication that non-Markovian dynamics can help a faster relaxation. Until this moment, all results showed that heating was faster than cooling, but it is known from literature that this asymmetry can be inverted. This motivated the investigation of how we could have cooling faster than heating in our systems. Using the knowledge that squeezed states relaxate faster than displaced states, we squeezed the hotter state and displaced the colder state. With this combination we inverted the asymmetry partially (for smaller parameters) and totally (for bigger parameters). We learned that the full inversion is generated only by the ergotropic contribution, while heating is still faster than cooling for the passive states. Also, we discussed that the partial inversion is actually a consequence of squeezing having a faster relaxation than displaced states, which was already discussed in the ergotropic Mpemba effect \cite{medina-2025}. In this case, the initial entropy production failed to indicate the faster process, but its time evolution helped understand the partial inversion. 

Beyond the investigations done in this work, there is a variety of quantum properties and systems where heating-cooling asymmetry could be generated. Regarding the model of two-interacting qubits, there is also the possibility of investigating other interaction hamiltonians, such as XXZ, XYZ, and transverse field Ising. These interaction models have anisotropies that may affect the heating-cooling asymmetry, for example, creating an inversion. Another interesting investigation is regarding quantum correlations in systems with more than two levels. It is known that for two-level systems heating is always faster than cooling \cite{vu-2021}, so it makes sense that global coherence and entanglement have not inverted the asymmetry. However, for systems with more than two levels, the asymmetry can be inverted depending on the energy gaps and degeneracies \cite{vu-2021}, which brings the question if quantum correlations could also cause the inversion for higher dimension systems. In this work, we also showed that non-Markovianity may accelerate the relaxation dynamics. In this respect, one possible future investigation would be to quantify non-Markovianity to verify if there is a linear correspondence of more information feedback and faster relaxation.


In the context of Gaussian states, we can also investigate how quantum resources, such as coherence, affect the heating-cooling asymmetry. Despite of non-zero ergotropy being an indicator of coherence in Gaussian states, this does not imply that ergotropy can quantify it. Using a proper quantifier can allow us to separate what of the observed behaviour in the relaxation of displaced and squeezed states is actually a consequence of quantum properties. Another interesting extension of our analysis would be investigating the consequences of entanglement in multi-mode Gaussian states. In this case, we have the two-mode squeezing operator that will change the ergotropic contribution of the Wigner-Fisher information, resulting in another effect to investigate in the relaxation process. As we briefly discussed in Sec. \ref{sec:result-squeezed}, the passive state of squeezed states can have a non-Markovian dynamics that seems to accelerate the relaxation. Therefore, in addition to the future direction for the discrete case, we can also investigate the relation of non-Markovianity and faster relaxation in Gaussian systems. Another possible future direction would be to study why entropy production is asymmetric. Onsager relations shows that entropy production can be written in terms of currents and  Ref. \cite{landi-2020} showed that it becomes non-linear and a not even function for arbitrarily far-from-equilibrium Gaussian states. A consequence of being a not even function is that the entropy production depends on the direction of the flow, which could be used to explain why heating and cooling are asymmetric in Gaussian systems.


