%% USPSC-Abstract.tex
%\autor{Silva, M. J.}
\begin{resumo}[Abstract]
 \begin{otherlanguage*}{english}
	\begin{flushleft} 
		\setlength{\absparsep}{0pt} % ajusta o espaçamento dos parágrafos do resumo		
 		\SingleSpacing  		\imprimirautorabr~~\textbf{\imprimirtitleabstract}.	\imprimirdata.  \pageref{LastPage} p. 
		%Substitua p. por f. quando utilizar oneside em \documentclass
		%\pageref{LastPage} f.
		\imprimirtipotrabalhoabs~-~\imprimirinstituicao, \imprimirlocal, 	\imprimirdata. 
 	\end{flushleft}
	\OnehalfSpacing 
It is known that classical thermodynamics fails to explain the thermal relaxation of far-from-equilibrium systems. Recent studies have reported an asymmetry in this regime, where heating is faster than cooling for thermodynamically equidistant initial states. Understanding how quantum systems equilibrate is a fundamental question with implications for quantum technologies, such as state preparation for quantum computing. However, how quantum resources affect this asymmetry is an important open question. In this work, we investigate the heating and cooling asymmetry in discrete and continuous-variable quantum systems. For the discrete case, we study two interacting qubits coupled to a global bosonic bath, focusing on the effects of initial correlations and local non-Markovian dynamics. For the continuous case, we consider Gaussian states and compare the thermal relaxation of displaced and squeezed states. Our results for the discrete case indicate that non-Markovianity may accelerate thermal relaxation, while the initial coherences do not change the equilibration time. In Gaussian systems, applying the displacement operator in the initial states does not affect the asymmetry, but using initially squeezed states can reduce it.

   \vspace{\onelineskip}
 
   \noindent 
   \textbf{Keywords}: Asymmetry, Relaxation, Two-Interacting Qubits, Gaussian Systems 
 \end{otherlanguage*}
\end{resumo}
